---
title: "Panel product"
output:
  pdf_document:
    highlight: pygments
vignette: >
  %\VignetteIndexEntry{panel_prod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DCM)
library(apollo)
```


Taking the product over observations for each individual (i.e., the probability of observing a sequence of choices) makes only a difference if we have random effects (because there we need the same draw for an individual for all observations).

Without the REC there is essentially no difference between the logit model with panel data and with purely cross-sectional data (Train, p. 55).


```{r}
apollo_initialise()

apollo_control = list(
  modelName       = "MNL_SP",
  modelDescr      = "Simple MNL model on mode choice SP data",
  indivID         = "ID",
  outputDirectory = "bla"
)

database = apollo_modeChoiceData

database = subset(database,database$SP==1)

apollo_beta=c(asc_car      = 0,
              asc_bus      = 0,
              asc_air      = 0,
              asc_rail     = 0,
              b_tt_car     = 0,
              b_tt_bus     = 0,
              b_tt_air     = 0,
              b_tt_rail    = 0,
              b_access     = 0,
              b_cost       = 0,
              b_no_frills  = 0,
              b_wifi       = 0,
              b_food       = 0)

apollo_fixed = c("asc_car","b_no_frills")

apollo_inputs = apollo_validateInputs()

apollo_probabilities=function(apollo_beta, apollo_inputs, functionality="estimate"){
  
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))
  
  P = list()
  
  V = list()
  V[["car"]]  = asc_car  + b_tt_car  * time_car                           + b_cost * cost_car
  V[["bus"]]  = asc_bus  + b_tt_bus  * time_bus  + b_access * access_bus  + b_cost * cost_bus 
  V[["air"]]  = asc_air  + b_tt_air  * time_air  + b_access * access_air  + b_cost * cost_air    + b_no_frills * ( service_air == 1 )  + b_wifi * ( service_air == 2 )  + b_food * ( service_air == 3 )
  V[["rail"]] = asc_rail + b_tt_rail * time_rail + b_access * access_rail + b_cost * cost_rail   + b_no_frills * ( service_rail == 1 ) + b_wifi * ( service_rail == 2 ) + b_food * ( service_rail == 3 )

  mnl_settings = list(
    alternatives  = c(car=1, bus=2, air=3, rail=4), 
    avail         = list(car=av_car, bus=av_bus, air=av_air, rail=av_rail), 
    choiceVar     = choice,
    utilities     = V
  )
  
  P[["model"]] = apollo_mnl(mnl_settings, functionality)
  
  P = apollo_panelProd(P, apollo_inputs, functionality)

  P = apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

out_1 <- apollo_probabilities(apollo_beta, apollo_inputs)
model_1 = apollo_estimate(apollo_beta, apollo_fixed, apollo_probabilities, apollo_inputs)
```


No `apollo_panelProd()`

```{r}
apollo_probabilities=function(apollo_beta, apollo_inputs, functionality="estimate"){
  
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))
  
  P = list()
  
  V = list()
  V[["car"]]  = asc_car  + b_tt_car  * time_car                           + b_cost * cost_car
  V[["bus"]]  = asc_bus  + b_tt_bus  * time_bus  + b_access * access_bus  + b_cost * cost_bus 
  V[["air"]]  = asc_air  + b_tt_air  * time_air  + b_access * access_air  + b_cost * cost_air    + b_no_frills * ( service_air == 1 )  + b_wifi * ( service_air == 2 )  + b_food * ( service_air == 3 )
  V[["rail"]] = asc_rail + b_tt_rail * time_rail + b_access * access_rail + b_cost * cost_rail   + b_no_frills * ( service_rail == 1 ) + b_wifi * ( service_rail == 2 ) + b_food * ( service_rail == 3 )
  
  mnl_settings = list(
    alternatives  = c(car=1, bus=2, air=3, rail=4), 
    avail         = list(car=av_car, bus=av_bus, air=av_air, rail=av_rail), 
    choiceVar     = choice,
    utilities     = V
  )
  
  P[["model"]] = apollo_mnl(mnl_settings, functionality)
  
  # P = apollo_panelProd(P, apollo_inputs, functionality)
  
  P = apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

database$ID <- 1:nrow(database)
apollo_inputs = apollo_validateInputs()

out_2 <- apollo_probabilities(apollo_beta, apollo_inputs)
model_2 = apollo_estimate(apollo_beta, apollo_fixed, apollo_probabilities, apollo_inputs)
```

The standard errors (derived from the Hessian) are not exactly the same. Are these numerical differences?

```{r}
apollo_modelOutput(model_1)
apollo_modelOutput(model_2)
```


